---
title: "Multiple-Discrete Choice Extreme Value Models with Continuous and Discrete Individual Heterogeneity in R: The rmdcev Package."
author:
  name: Patrick Lloyd-Smith
  affiliation: University of Saskatchewan
  # email: grantmcd@uoregon.edu
date: April 2019  #"`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    theme: flatly
    highlight: haddock 
    # code_folding: show
    toc: yes
    toc_depth: 4
    toc_float: yes
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, dpi=300)
```

This paper introduces the package **rmdcev** in R for estimation of multiple-discrete choice extreme value models with unobserved heterogeneity across individuals as well as subsequent demand and welfare simulation. 

The models supported by rmdcev are multiple-discrete choice extreme value, the latent class, the mixed- multiple-discrete choice extreme value
The models are estimated using either penalized maximum likelihood estimator or hierarchal Bayes. 
Demand and welfare simulations can 
This article describes and demonstrates with real and simulated datasets all the functionalities of rmdcev.

 for cross-sectional and panel (longitudinal). The package also simulates demand and welfare implications. 
The purpose of this paper is to describe the model

Keywords: multiple-discrete choice extreme value, latent class, random parameters, welfare, preference heterogeneity, R.

## Introduction

Individual choice contexts are often characterized by both extensive (i.e. what alternative to choose) and intensive (i.e. how much of an alternative to consume) margins. These multiple discrete-continuous (MDC) choice situations are pervasive, arising in transportation, marketing, and decisions regarding environmental resources.
Kuhn-Tucker (KT) consumer demand models are often employed to analyze these MDC situations and substantial progress has been made on improving the econometric modeling structures. There is currently no available software package that 


One reason cited for the lack of widespread use of these KT models is that applying these models for welfare analysis is not straightforward \citep{von_haefen_kuhn-tucker_2005, bhat_multiple_2014}. This issue is especially relevant in applying these models to studying decisions regarding environmental resources where producing welfare estimates is often the main purpose of the research.

I present a brief example below showing

## Models

### Conceptual framework

Each individual $i$ maximizes utility through the choice of the numeraire good ($\x_1$) and the non-numeraire alternatives ($x_k$) subject to a monetary budget constraint. We assume there is a numeraire good (i.e. essential Hicksian composite good) which is always consumed and has a price of one. The individual's maximization problem is

\begin{equation}
\max_{x_k, x_1} U(x_k, x_1) \; \;  \; \; s.t. \; \; y = \sum\limits^K_{k=2}p_k x_k + x_1,x_k \geq 0, k = 2,...,K  
\end{equation}

\noindent
where $x_k$ is the consumption level for alternative $k$, $x_1$ is consumption of the numeraire, $y$ is annual income, and $p_k$ is the unit price of alternative $k$.

The resulting first-order KT conditions that implicitly define the solution to the optimal consumption bundles of $x_k$ and $x_1$ are

\begin{eqnarray}
 \frac{U_{x_k}}{U_{x_1}} &\leq p_k,\; \; k = 1,....K,  \\
  x_k\left[\frac{U_{x_k}}{U_{x_1}} - p_k \right] &= 0,\; \; k = 1,....K.
\label{ktconditions}
\end{eqnarray}

For alternatives with positive consumption levels, the marginal rate of substitution between these alternatives and the numeraire good is equal to the price of the alternative. For unconsumed alternatives, the marginal rate of substitution between these alternatives and the numeraire good is less than the price of the alternatives.

## MDCEV

The **rmdcev** pacakge implements the random utility specification as introduced by \cite{bhat2008multiple}. The utility function is specified as

\begin{equation}
\label{utilkt}
U(x_k, x_1) = \sum_{k=2}^{K} \frac{\gamma_k}{\alpha_k}\psi_k \left[ \left( \frac{x_k}{\gamma_k} + 1 \right)^{\alpha_k} - 1 \right] + \frac{\psi_1}{\alpha_1}x_1^{\alpha_1}
\end{equation}

\noindent
where $x_k$ is the consumption level for alternative $k$, $x_1$ is consumption of the numeraire, $y$ is annual income, and $p_k$ is the unit price of alternative $k$.

where $\gamma_k, \psi_k > 0$ and $\alpha_k, \alpha_1 \leq 1$ for all $k$ are required for this function to be consistent with the properties of a utility function \citep{bhat2008multiple}. Bhat (2008) provides an overview of the parameter interpretation and in brief

- The $\psi_k$ parameters represent the marginal utility of consuming alternative $k$ at the point of zero consumption (i.e. baseline marginal utility).
- The $\gamma_k$ parameters are translation parameters that influence satiation and allow for corner solutions (i.e. zero consumption levels for alternatives). The lower the value of $\gamma_k$, the more satiation effect in consuming $x_k$.
- The $\alpha$ parameters control the rate of diminishing marginal utility of additional consumption. If $\alpha_k$ equal to one, then there is no satiation effects (i.e. constant marginal utility). 

A multiplicative random element is introduced into the baseline marginal utility as 

\begin{equation}
\psi_k=\psi(z_k,\varepsilon_k)= exp(\beta'z_k+\varepsilon_k)
\end{equation}

\noindent
where $\z_k$ is a set of variables that can include alternative-specific attributes and individual-specific characteristics, and $\varepsilon_k$ is an error term that allows for the utility function to be random over the population.

We assume an extreme value distribution that is independently distributed across alternatives for $\varepsilon_k$ with an associated scale parameter of $\sigma$.

We specify the following parameterization of the utility function parameters

$\psi_k=exp(\beta_q Q_k + \beta_s S_i + \varepsilon_k)$, 

$\gamma_k = exp(\gamma_k)$, and

$\alpha = 1 - exp(\alpha)$.

For identification, specify $\psi_1= e^{\varepsilon_1}$.

Weak complementarity, which is required for deriving unique welfare measures, is imposed in this specification by adding and subtracting one in the non-numeraire part of the utility function.


Bhat (2008) discusses the identification concerns regarding estimating separate $\gamma_k$ and $\alpha_k$ parameters for each non-numeraire alternative. As such, there are three possible utility function specifications

1. $\alpha$-profle:

\begin{equation}
\label{utilkt_alpha}
U(x_k, x_1) = \sum_{k=2}^{K} \frac{1}{\alpha_k}exp(\beta'z_k+\varepsilon_k) \left[ \left( x_k + 1 \right)^{\alpha_k} - 1 \right] + \frac{exp(\varepsilon_1)}{\alpha_1}x_1^{\alpha_1}
\end{equation}

2. $\gamma$-profle:

\begin{equation}
\label{utilkt_gamma}
U(x_k, x_1) = \sum_{k=2}^{K} \gamma_k exp(\beta'z_k+\varepsilon_k) \ln\left( \frac{x_k}{\gamma_k} + 1 \right) + \frac{exp(\varepsilon_1)}{\alpha_1}x_1^{\alpha_1}
\end{equation}

3. Hybrid-profile:

\begin{equation}
\label{utilkt_hybrid}
U(x_k, x_1) = \sum_{k=2}^{K} \frac{\gamma_k}{\alpha} exp(\beta'z_k+\varepsilon_k) \left[ \left( \frac{x_k}{\gamma_k} + 1 \right)^{\alpha_k} - 1 \right] + \frac{exp(\varepsilon_1)}{\alpha}x_1^{\alpha}
\end{equation}


The resulting model probability of the consumption pattern where $M$ goods are chosen can be expressed as \citep{bhat2008multiple}

\begin{equation}
P(x^{*}_1,x^{*}_2...x^{*}_M,0,...,0) = \frac{1}{\sigma^{M-1}} \left(\prod_{m=1}^M c_m \right)\left(\sum_{m=1}^M \frac{1}{c_j} \right) \left( \ \frac{\prod_{m=1}^M e^{V_m/\sigma}}{ \left( \sum_{k=1}^J e^{V_k/\sigma} \right)^M }\right)(M-1)!
\end{equation}

\noindent
where $\sigma$ is a scale parameter and $c_m = \frac{1-\alpha}{x_m+ \gamma_m}$. The $V$ expressions depend on what model specification is used:

1. $\alpha$-profle: $V_k = \beta' z_k + (\alpha_k-1)\ln\left( x_k + 1 \right) - \ln \left(p_k\right)$ for $k \geq 2$, and $V_1 = (\alpha_1-1)\ln(x_1)$. 

2. $\gamma$-profle: $V_k = \beta' z_k - \ln\left( \frac{x_k}{\gamma_k} + 1 \right) - \ln \left(p_k\right)$ for $k \geq 2$, and $V_1 = (\alpha_1-1)\ln(x_1)$. 

3. Hybrid-profile: $V_k = \beta' z_k + (\alpha-1)\ln\left( \frac{x_k}{\gamma_k} + 1 \right) - \ln \left(p_k\right)$, and $V_1 = (\alpha-1)\ln(x_1)$. 

## Latent Class MDCEV

## Random parameters

## The rmdcev package

### Data format

The rmdcev pacakage accepts data in ``long" format (one row per available non-numeraire alternative). The following columns are needed:

- **id**: The individual level id variable
- **good**: The variable containing the good number
- **quant**: Consumption levels for each good
- **price**: Price levels for each good
- **inc**: Income level for each individual

Data must be arragned by id and then good. 

```{r, cache=F, message=F}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, rstan, rmdcev, hrbrthemes)
theme_set(hrbrthemes::theme_ipsum())
```
```{r cache=F, include=F}
## Switch off automatic "pretty printing" of DFs
## See: https://hughjonesd.github.io/huxtable/huxtable.html#automatic-pretty-printing-of-data-frames
options(huxtable.knit_print_df = FALSE)
```

```{r, data}
data(recreation, package = "rmdcev")
head(data, n = 20)
```

We can summarize the average consumption and price levels for each good as:

```{r, summary}
data %>%
	group_by(good, activity) %>%
	summarise(mean_quant = mean(quant),
			  mean_price = mean(price))
```



###	Formula interface

The $z_k$ variables can be specified using the **psi_formula** interface. This formula is based on the R package **Formula** (Zeileis and Crossiant, 2010).

```r
psi_formula = ~ b1 + b2 + b3 + b4 + b5 + b6 + b7 + b8-1
```

Note that the formula will automatically include a constant but a constant can be omitted if -1 is used in the formula. 



### Estimating MDCEV 

We estimate a MDCEV model using the **Recreation** data where we include acticity-specific constants in the $\psi$ parameter. We start with the maximum likelihood estimation by setting algorithm = "MLE" and using the gamma model specification where alpha is set to 0. 

The basic syntax is the following:

```{r, estimation_mdcev}
library(rmdcev)
mdcev.fit <- FitMDCEV(psi_formula = ~ factor(activity) -1,
							data = data,
							model = "gamma0",
							algorithm = "MLE")
```

There are four possible model options 

- "les": Linear expenditure system with one estimated $\alpha_1$ and all non-numeraire $\alpha$ parameters equal to 1.
- "alpha": $\alpha$-profile with all $\gamma_k$ equal to 1. 
- "gamma": $\gamma$-profile with a single estimated $\alpha$ (i.e. $\alpha_1 = \alpha_k = \alpha$. 
- "gamma0": $\gamma$-profile with all  $\alpha$ equal to 1e-6.

### Estimating MDCEV with HB

### Estimating LC_MDCEV

In this example, we estimate a LC model with two classes. To include variables in the membership equation we specify them in the **lc_formula** interface. The LC model is estimated as long as **n_classes** is set greater than 2. 

```{r, estimation_mdcev_lc}
library(rmdcev)
mdcev.fit <- FitMDCEV(psi_formula = ~ factor(activity) -1,
					  lc_formula = ~ university,
							data = data,
					  		n_classes = 2,
							model = "gamma0",
							algorithm = "MLE")
```

### Estimating Mixed MDCEV

*too come*

## Rmdcev Demand and Welfare Simulation

The **rmdcev** package includes 

### Unconditional vs Conditional Error Draws


These first-order conditions can be used to derive Hicksian demands and welfare measures. The Hicksian compensating surplus ($CS^H$) for a change in price and quality from baseline levels $p^0$ and $q^0$ to new `policy' levels $p^1$ and $q^1$ is defined implicitly using an indirect utility function

\begin{align}
V(p^0, q^0, y, \theta, \varepsilon) = V(p^1, q^1, y - CS^H, \theta, \varepsilon)
\label{welfare_indirect} 
\end{align}

\noindent
or explicitly using an expenditure function

\begin{align}
CS^H = y - e(p^1, q^1, \bar{U}, \theta, \varepsilon)
\label{welfare} 
\end{align}

\noindent
where $y$ is income, $\theta$ is the vector of structural parameters ($\psi_k, \alpha_k, \alpha_1, \gamma_k$), $\varepsilon$ is a vector or matrix of unobserved heterogeneity, and $\bar{U} = V(p^0, q^0, y, \theta, \varepsilon)$. Before describing the new approach to welfare measurement, I first review main difficulties in welfare measurement in KT models and describe the existing approaches employed in the literature.

### Specific steps in algorithm for Hybrid-profile utility functions

**Step 0**: Assume that only the numeraire alternative is chosen and let the number of chosen alternatives equal one (M=1).

**Step 1**: Using the data, model parameters, and either conditional or unconditional simulated error term draws, calculate the price-normalized baseline utility values ($\psi_k/p_k$) for all alternatives. Sort the $K$ alternatives in the descending order of their price-normalized baseline utility values. Note that the numeraire alternative is in the first place. Go to step 2.

**Step 2**: Compute the value of $\lambda^E$ using the following equation:

\begin{align}
\label{lambdaE}
\frac{1}{\lambda^E} = \left[ \frac{\alpha \bar{U} + \sum_{m=2}^{M} \gamma_m \psi_m} {\sum_{m=2}^{M} \gamma_m \psi_m \left( \frac{p_m}{\psi_m} \right)^\frac{\alpha}{\alpha-1} + \psi_1 \left(\frac{p_1}{\psi_1} \right)^\frac{\alpha}{\alpha-1}} \right] ^\frac{\alpha-1}{\alpha}.
\end{align}

Go to step 3.

**Step 3**: If $\frac{1}{\lambda^E} > \frac{\psi_{M+1}}{p_{M+1}}$, go to step 4. Else if $\frac{1}{\lambda^E} < \frac{\psi_{M+1}}{p_{M+1}}$, set $M = M + 1$. If $M < K$, go back to step 2. If $M = K$, go to step 4. 

**Step 4**: Compute the optimal Hicksian consumption levels for the first $I$ alternatives in the above descending order using the following equations

\begin{align}
\begin{split}
\label{foc13}
x_1 &=   \left( \frac{p_1}{\lambda^E \psi_1} \right)^\frac{1}{\alpha_1-1}\text{, and} \\
x_m &=   \left[ \left( \frac{p_m}{\lambda^E \psi_m} \right)^\frac{1}{\alpha_m-1}-1 \right]\gamma_m, \; \; \text{if} \; \; x_m > 0. 
\end{split}
\end{align}

Set the remaining alternative consumption levels to zero and stop. 

### Specific steps in algorithm for general utility functions:

In the more general case of a utility function with different $\alpha_k$ parameters for each alternative, I modify the algorithm in Section 3.2. In this context, there is no closed-form expressions for $\lambda^E$ and I need to conduct a numerical bisection routine. Let $\hat{\lambda^E}$ and $\hat{U}$ be estimates of $\lambda^E$ and $U$ and let $tol_{\lambda}$ and $tol_{U}$ be the tolerance levels for estimating $\lambda^E$ and $U$ that can be arbitrarily small.

**Step 0**: Assume that only the numeraire is chosen and let the number of chosen alternatives equal one (M=1).

**Step 1**: Using the data, model parameters, and either conditional or unconditional simulated error term draws, calculate the price-normalized baseline utility values ($\psi_k/p_k$) for all alternatives. Sort the $K$ alternatives in the descending order of their price-normalized baseline utility values. Note that the numeraire is in the first place. Go to step 2.

**Step 2**: Let $\frac{1}{\hat{\lambda^E}} = \frac{\psi_{M+1}}{p_{M+1}}$ and substitute $\hat{\lambda^E}$ into the following quation to obtain an estimate of $\hat{U}$.

\begin{align}
\bar{U}=\sum_{M=2}^{M} \frac{\gamma_m}{\alpha_m}\psi_m \left[ \left( \frac{p_m}{\lambda^E \psi_m} \right)^\frac{\alpha_m}{\alpha_m-1} - 1 \right] + \frac{\psi_1}{\alpha_1}\left(\frac{p_1}{\lambda^E \psi_1} \right)^\frac{\alpha_1}{\alpha_1-1}.
\label{util_algo}
\end{align}

**Step 3**: If $\hat{U} < \bar{U}$, go to step 4. Else, if $\hat{U} \geq \bar{U}$, set $\frac{1}{\lambda_l^E}= \frac{\psi_{M+1}}{p_{M+1}}$ and $\frac{1}{\lambda_u^E}= \frac{\psi_{M}}{p_{M}}$. Go to step 5.

**Step 4**: Set $M=M+1$. If $M<K$, go to step 2. Else if $M=K$, set $\frac{1}{\lambda_l^E}= 0$ and $\frac{1}{\lambda_u^E}= \frac{\psi_{K}}{p_{K}}$. Go to step 5.

**Step 5**: Let $\hat{\lambda^E}= (\lambda_l^E+\lambda_u^E)/2$ and substitute $\hat{\lambda^E}$ into Equation (\ref{util_algo}) to obtain an estimate of $\hat{U}$. Go to step 6.

**Step 6**: If $|\lambda_l^E-\lambda_u^E| \leq \; tol_{\lambda}$ or $|\hat{U}-\bar{U}| \leq \; tol_{U}$, go to step 7. Else if $\hat{U}<\bar{U}$, update $\lambda^E_u= (\lambda_l^E+\lambda_u^E)/2$ and go to step 5. Else if $\hat{U}>\bar{U}$, update $\lambda^E_l= (\lambda_l^E+\lambda_u^E)/2$ and go to step 5.

**Step 7**: Compute the optimal Hicksian consumption levels for the first $I$ alternatives in the above descending order using Equation (\ref{foc13}). Set the remaining alternative consumption levels to zero and stop. 

## Modelling notes



## Conclusions

The **rmdcev** package implements the MDCEV . In this paper, we have demonstrated the use of the pacakage to estimate several model specifications and to derive welfare.
