---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Multiple Discrete-Continuous Extreme Value (MDCEV) Model Estimation and Simulation in R: The rmdcev Package 

<!-- badges: start -->
[![Build Status](https://travis-ci.org/plloydsmith/rmdcev.svg?branch=master)](https://travis-ci.org/plloydsmith/rmdcev)
<!-- badges: end -->

The _rmdcev_ R package estimates and simulates multiple discrete-continuous extreme value (MDCEV) demand models with observed and unobserved individual heterogneity. Fixed parameter, latent class, and random parameter models can be estimated. These models are estimated using maximum likelihood or Bayesian estimation techniques and are implemented in Stan, which is a C++ package for performing full Bayesian inference (see http://mc-stan.org/). The package also includes functions for simulating demand and welfare outcomes from policy scenarios.

## Current Status

Development is in progress. Currently users can estimate the following models:

1. Fixed parameter models (maximum likelihood or Bayesian estimation)
2. Latent class models (maximum likelihood estimation)
3. Random parameters models (Bayesian estimation)

Under development

2. Phaneuf/von Haefen Kuhn Tucker model

## Installation

I recommend you first install **rstan** and C++ toolchain by following these steps:

https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started

Once **rstan** is installed, you can install the released version of rmdcev from GitHub using devtools

``` r
if (!require(devtools)) {
  install.packages("devtools")
  library(devtools)
}
install_github("plloydsmith/rmdcev", build_vignettes = FALSE)
```

Depending on your computer set-up, you may need to adjust your Makevars file of the .R folder (usually in your computer user folder) to ensure the first two lines are

``` r
CXX14 = $(BINPREF)g++ -m$(WIN) -std=c++1y
CXX14FLAGS=-O3 -mtune=native -march=native
```

You can switch build_vignettes to TRUE but it will take a lot longer to install. If installation fails, please let me know by [filing an issue](https://github.com/plloydsmith/rmdcev/issues).

## References

For more details on the model specification and estimation:

Bhat, C.R. (2008) ["The Multiple Discrete-Continuous Extreme Value (MDCEV) Model: Role of Utility Function Parameters, Identification Considerations, and Model Extensions."](https://www.sciencedirect.com/science/article/pii/S0191261507000677) Transportation Research Part B, 42(3): 274-303.

For more details on the welfare simulation:

Lloyd-Smith, P (2018). ["A New Approach to Calculating Welfare Measures in Kuhn-Tucker Demand Models."](https://www.sciencedirect.com/science/article/pii/S1755534517300994) Journal of Choice Modeling, 26: 19-27

## Estimation

As an example, we can simulate some data using the Hybrid specification. In this example, we are simulating data for 1000 invdividuals and 10 non-numeraire alternatives.

```{r}
library(pacman)
p_load(tidyverse, rmdcev)
model <- "hybrid"
nobs <- 1000
ngoods <- 10
sim.data <- GenerateMDCEVData(model = model, nobs = nobs, ngoods = ngoods)
```

Estimate model using MLE
``` {r}
mdcev_est <- FitMDCEV(psi_formula = ~ b1 + b2 + b3 + b4 + b5 + b6 + b7 + b8-1,
				   data = sim.data$data,
				   model = model,
				   algorithm = "MLE")
```

Summarize results
``` {r}
SummaryMDCEV(mdcev_est)
```


Compare estimates to true values
``` {r}
parms_true <- tbl_df(sim.data$parms_true) %>%
	mutate(true = as.numeric(true))

coefs <- mdcev_est$est_pars %>%
	mutate(parms = gsub("\\[|\\]", "", parms)) %>%
	group_by(parms) %>%
	summarise(mean = mean(value),
			  sd = sd(value),
			  zstat = mean / sd,
			  cl_lo = quantile(value, 0.025),
			  cl_hi = quantile(value, 0.975)) %>%
	left_join(parms_true, by = "parms") %>%
	print(n=200)
```

Compare outputs using a figure
```{r, warning = FALSE}
coefs %>%
	ggplot(aes(y = mean, x = true))  +
	geom_point(size=2) +
	geom_text(label=coefs$parms) +
	geom_abline(slope = 1) +
	geom_errorbar(aes(ymin=cl_lo,ymax=cl_hi,width=0.2))
```


## Welfare simulation

Create policy simulations (these are 'no change' policies with no effects)
```{r}
npols <- 2 # Choose number of policies

policies<-	CreateBlankPolicies(npols, ngoods, mdcev_est$stan_data[["dat_psi"]], price_change_only = TRUE)

df_sim <- PrepareSimulationData(mdcev_est, policies)
```

Simulate welfare changes
```{r}
wtp <- SimulateMDCEV(df_sim$df_indiv, df_common = df_sim$df_common, sim_options = df_sim$sim_options,
					 cond_err = 1, nerrs = 5, sim_type = "welfare")
SummaryWelfare(wtp)
```
