---
title: "Multiple-Discrete Choice Extreme Value Model Estimation and Simulation in R: The rmdcev Package"
author:
  name: Patrick Lloyd-Smith
  affiliation: University of Saskatchewan
date: April 2019  #"`r format(Sys.time(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple-Discrete Choice Extreme Value Model Estimation and Simulation in R: The rmdcev Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
#  html_document:
#    theme: flatly
#    highlight: haddock
#    toc: yes
#    toc_depth: 4
#    toc_float: yes
#    keep_md: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, dpi=300)
library(rmdcev)
```

## Abstract

This paper introduces the package **rmdcev** in R for estimation and simulation of multiple-discrete choice extreme value models with individual heterogeneity. The models supported by **rmdcev** are the multiple-discrete choice extreme value (MDCEV) model, the latent class MDCEV model, and the random parameters MDCEV model. The models are fit using maximum likelihood estimation or Bayesian estimation. The **rmdcev** package also implements demand forecasting and welfare calculation for policy simulation. The purpose of this paper is to describe the MDCEV estimation and simulation framework and demonstrates all the functionalities of **rmdcev**. 

Keywords: multiple-discrete choice extreme value, Kuhn-Tucker, latent class, random parameters, demand, welfare, preference heterogeneity, R.

## Introduction

Individual choice contexts are often characterized by both extensive (i.e. what alternative to choose) and intensive (i.e. how much of an alternative to consume) margins. These multiple discrete-continuous (MDC) choice situations are pervasive, arising in transportation, marketing, health, and decisions regarding environmental resources.
Kuhn-Tucker (KT) consumer demand models are often employed to analyze these MDC situations and substantial progress has been made on improving the econometric modeling structures. There is currently no available software package that

In addition to **rmdcev**, the **Apollo** package from the Leeds Choice Modeling Centre provides a flexible interface to estimating MDCEV models and simulating demand. The main differences is that **rmdcev** 1) provides welfare simulation, 2) also implements the Kuhn-Tucker Linear Expenditure System utility specification, and 3) uses Stan software and NUTs sampling.


One reason cited for the lack of widespread use of these KT models is that applying these models for welfare analysis is not straightforward \citep{von_haefen_kuhn-tucker_2005, bhat_multiple_2014}. This issue is especially relevant in applying these models to studying decisions regarding environmental resources where producing welfare estimates is often the main purpose of the research.

I present a brief example below showing

## Models

### Conceptual framework

Each individual $i$ maximizes utility through the choice of the numeraire good ($\x_1$) and the non-numeraire alternatives ($x_k$) subject to a monetary budget constraint. We assume there is a numeraire good (i.e. essential Hicksian composite good) which is always consumed and has a price of one. The individual's maximization problem is

\begin{equation}
\max_{x_k, x_1} U(x_k, x_1) \; \;  \; \; s.t. \; \; y = \sum\limits^K_{k=2}p_k x_k + x_1,x_k \geq 0, k = 2,...,K  
\end{equation}

\noindent
where $x_k$ is the consumption level for alternative $k$, $x_1$ is consumption of the numeraire, $y$ is annual income, and $p_k$ is the unit price of alternative $k$.

The resulting first-order KT conditions that implicitly define the solution to the optimal consumption bundles of $x_k$ and $x_1$ are

\begin{eqnarray}
 \frac{U_{x_k}}{U_{x_1}} &\leq p_k,\; \; k = 1,....K,  \\
  x_k\left[\frac{U_{x_k}}{U_{x_1}} - p_k \right] &= 0,\; \; k = 1,....K.
\label{ktconditions}
\end{eqnarray}

For alternatives with positive consumption levels, the marginal rate of substitution between these alternatives and the numeraire good is equal to the price of the alternative. For unconsumed alternatives, the marginal rate of substitution between these alternatives and the numeraire good is less than the price of the alternatives.

These first-order conditions can be used to derive Hicksian demands and welfare measures. The Hicksian compensating surplus ($CS^H$) for a change in price and quality from baseline levels $p^0$ and $q^0$ to new `policy' levels $p^1$ and $q^1$ is defined implicitly using an indirect utility function

\begin{align}
V(p^0, q^0, y, \theta, \varepsilon) = V(p^1, q^1, y - CS^H, \theta, \varepsilon)
\label{welfare_indirect}
\end{align}

\noindent
or explicitly using an expenditure function

\begin{align}
CS^H = y - e(p^1, q^1, \bar{U}, \theta, \varepsilon)
\label{welfare}
\end{align}

\noindent
where $y$ is income, $\theta$ is the vector of structural parameters ($\psi_k, \alpha_k, \gamma_k$), $\varepsilon$ is a vector or matrix of unobserved heterogeneity, and $\bar{U} = V(p^0, q^0, y, \theta, \varepsilon)$.

## MDCEV

The **rmdcev** package implements the random utility specification as introduced by \cite{bhat2008multiple}. The utility function is specified as

\begin{equation}
\label{utilkt}
U(x_k, x_1) = \sum_{k=2}^{K} \frac{\gamma_k}{\alpha_k}\psi_k \left[ \left( \frac{x_k}{\gamma_k} + 1 \right)^{\alpha_k} - 1 \right] + \frac{\psi_1}{\alpha_1}x_1^{\alpha_1}
\end{equation}

\noindent
where $x_k$ is the consumption level for alternative $k$, $x_1$ is consumption of the numeraire, $y$ is annual income, and $p_k$ is the unit price of alternative $k$.

where $\gamma_k, \psi_k > 0$ and $\alpha_k, \alpha_1 \leq 1$ for all $k$ are required for this function to be consistent with the properties of a utility function \citep{bhat2008multiple}. Bhat (2008) provides an overview of the parameter interpretation and in brief

- The $\psi_k$ parameters represent the marginal utility of consuming alternative $k$ at the point of zero consumption (i.e. baseline marginal utility).
- The $\gamma_k$ parameters are translation parameters that allow for corner solutions (i.e. zero consumption levels for alternatives) and also influence satiation. The lower the value of $\gamma_k$, the more satiation effect in consuming $x_k$.
- The $\alpha$ parameters control the rate of diminishing marginal utility of additional consumption. If $\alpha_k$ equal to one, then there is no satiation effects (i.e. constant marginal utility).

A multiplicative random element is introduced into the baseline marginal utility as

\begin{equation}
\psi_k=\psi(z_k,\varepsilon_k)= exp(\beta'z_k+\varepsilon_k)
\end{equation}

\noindent
where $\z_k$ is a set of variables that can include alternative-specific attributes and individual-specific characteristics, and $\varepsilon_k$ is an error term that allows for the utility function to be random over the population. We assume an extreme value distribution that is independently distributed across alternatives for $\varepsilon_k$ with an associated scale parameter of $\sigma$.

We specify the following parameterization of the utility function parameters

$\psi_k=exp(\beta_q Q_k + \beta_s S_i + \varepsilon_k)$,

$\gamma_k = exp(\gamma_k)$, and

$\alpha = 1 - exp(\alpha)$.

For identification, specify $\psi_1= e^{\varepsilon_1}$.

Weak complementarity, which is required for deriving unique welfare measures, is imposed in this specification by adding and subtracting one in the non-numeraire part of the utility function.


Bhat (2008) discusses the identification concerns regarding estimating separate $\gamma_k$ and $\alpha_k$ parameters for each non-numeraire alternative. There are three possible utility function specifications

1. $\alpha$-profle:

\begin{equation}
\label{utilkt_alpha}
U(x_k, x_1) = \sum_{k=2}^{K} \frac{1}{\alpha_k}exp(\beta'z_k+\varepsilon_k) \left[ \left( x_k + 1 \right)^{\alpha_k} - 1 \right] + \frac{exp(\varepsilon_1)}{\alpha_1}x_1^{\alpha_1}
\end{equation}

2. $\gamma$-profle:

\begin{equation}
\label{utilkt_gamma}
U(x_k, x_1) = \sum_{k=2}^{K} \gamma_k exp(\beta'z_k+\varepsilon_k) \ln\left( \frac{x_k}{\gamma_k} + 1 \right) + \frac{exp(\varepsilon_1)}{\alpha_1}x_1^{\alpha_1}
\end{equation}

3. Hybrid-profile:

\begin{equation}
\label{utilkt_hybrid}
U(x_k, x_1) = \sum_{k=2}^{K} \frac{\gamma_k}{\alpha} exp(\beta'z_k+\varepsilon_k) \left[ \left( \frac{x_k}{\gamma_k} + 1 \right)^{\alpha_k} - 1 \right] + \frac{exp(\varepsilon_1)}{\alpha}x_1^{\alpha}
\end{equation}


The resulting model probability of the consumption pattern where $M$ goods are chosen can be expressed as \citep{bhat2008multiple}

\begin{equation}
P(x^{*}_1,x^{*}_2...x^{*}_M,0,...,0) = \frac{1}{\sigma^{M-1}} \left(\prod_{m=1}^M c_m \right)\left(\sum_{m=1}^M \frac{1}{c_j} \right) \left( \ \frac{\prod_{m=1}^M e^{V_m/\sigma}}{ \left( \sum_{k=1}^J e^{V_k/\sigma} \right)^M }\right)(M-1)!
\end{equation}

\noindent
where $\sigma$ is a scale parameter and $c_m = \frac{1-\alpha}{x_m+ \gamma_m}$. The $V$ expressions depend on what model specification is used:

1. $\alpha$-profle: $V_k = \beta' z_k + (\alpha_k-1)\ln\left( x_k + 1 \right) - \ln \left(p_k\right)$ for $k \geq 2$, and $V_1 = (\alpha_1-1)\ln(x_1)$.

2. $\gamma$-profle: $V_k = \beta' z_k - \ln\left( \frac{x_k}{\gamma_k} + 1 \right) - \ln \left(p_k\right)$ for $k \geq 2$, and $V_1 = (\alpha_1-1)\ln(x_1)$.

3. Hybrid-profile: $V_k = \beta' z_k + (\alpha-1)\ln\left( \frac{x_k}{\gamma_k} + 1 \right) - \ln \left(p_k\right)$, and $V_1 = (\alpha-1)\ln(x_1)$.

## Latent Class MDCEV

## Random parameters

## The rmdcev package

### Data format

The **rmdcev** package accepts data in ``long" format (one row per available non-numeraire alternative). There is no row for the numeraire good. The following named columns are required in the data:

- **id**: The id variable for each individual
- **good**: The variable containing the good number
- **quant**: Consumption levels for each good
- **price**: Price levels for each good
- **inc**: Income level for each individual

Data must be arranged by id and then good.

As an example, we can load the recreation data incorporated with the package. This data is from the Value of Nature to Canadians Survey and includes 17 different recreation activities for 2,000 people.

```{r, data}
data(data_rec, package = "rmdcev")
data_rec
```

We can summarize the average consumption and price levels for each good as:

```{r, summary}
library(dplyr)
data_rec %>%
	group_by(good, good_name) %>%
	summarise(mean_quant = mean(quant),
			  mean_price = mean(price))
```



###	Formula interface

The $z_k$ variables can be specified using the **psi_formula** interface. This formula is based on the R package **Formula** (Zeileis and Crossiant, 2010).

```r
psi_formula = ~ z1 + z2 + z3
```

Note that the formula will automatically include a constant but a constant can be omitted if -1 is used in the formula.

### Estimating MDCEV

We estimate a MDCEV model using the **Recreation** data where we include activity-specific constants (ASCs) in the $\psi$ parameter. To create the ASC data, we can use the following code to create indicator variables. Note that we can also use the factor(good_name) command directly in the formula but the names are quite long.

```{r, create_dummies}
data_rec <- data_rec %>% 
  mutate(v = 1, yr = good_name) %>% 
  spread(yr, v, fill = 0) %>%
	arrange(id, good)
```

We specify the gamma model specification where alpha is fixed to 0 by setting model = "gamma0". We use maximum likelihood estimation by setting algorithm = "MLE".

The syntax for the model is the following:

```{r, estimation_mdcev}
mdcev.fit <- FitMDCEV(psi_formula = ~ beach + birding + camping + cycling + fish + garden + golf +
					 hiking + hunt_birds + hunt_large + hunt_trap + hunt_waterfowl + motor_land + 
					 	motor_water + photo + ski_cross + ski_down -1,
							data = data_rec,
							model = "gamma0",
							algorithm = "MLE",
					  		print_iterations = FALSE)
```

The function first checks the data is in the correct form and then estimates the model.

The output of the function can be accessed by calling SummaryMDCEV. 

```{r, summary_mdcev}
	SummaryMDCEV(mdcev.fit)
```

The summary includes overal information on the model and the parameter estimates. 

There are four possible model options

- "les": Linear expenditure system with one estimated $\alpha_1$ and all non-numeraire $\alpha$ parameters equal to 1.
- "alpha": $\alpha$-profile with all $\gamma_k$ equal to 1.
- "gamma": $\gamma$-profile with a single estimated $\alpha$ (i.e. $\alpha_1 = \alpha_k = \alpha$.
- "gamma0": $\gamma$-profile with all  $\alpha$ equal to 1e-6.

### Estimating MDCEV with Bayesian Approaches

The exact same model can be fit using Bayesian estimation by changing the model call to "Bayes".


```{r, estimation_mdcev_bayes}
mdcev_bayes <- FitMDCEV(psi_formula = ~ beach + birding + camping + cycling + fish + garden + golf +
					 hiking + hunt_birds + hunt_large + hunt_trap + hunt_waterfowl + motor_land + 
					 	motor_water + photo + ski_cross + ski_down -1,
							data = data_rec,
							model = "gamma0",
							algorithm = "Bayes",
						n_iterations = 10,
						n_chains = 1,
					  		print_iterations = FALSE)
```

The function first checks the data is in the correct form and then estimates the model.

The output of the function can be accessed by calling SummaryMDCEV. 

```{r, summary_mdcev_bayes}
	SummaryMDCEV(mdcev_bayes)
```

One benefit of using the Bayesian approach is that we can take advantage of the postestimation commands in **rstan**, **bayesplot**, and **shinystan**.

### Estimating Latent Class MDCEV Models

In this example, we estimate a LC model using the **Recreation** data. We set the number of classes equal to 2. We would like to include the *university*, *ageindex*, and *urban* in the membership equation and we include them in the **lc_formula** interface. Note that we need to include at least a constant in the formula. The LC model is estimated as long as the prespecified number of classes (**n_classes**) is set greater than 2.

```{r, estimation_mdcev_lc}
mdcev.lc <- FitMDCEV(psi_formula = ~ beach + birding + camping + cycling + fish + garden + golf + 
								 	hiking + hunt_birds + hunt_large + hunt_trap + hunt_waterfowl + 
								 	motor_land + motor_water + photo + ski_cross + ski_down - 1,
					lc_formula = ~ university + ageindex + urban,
					data = data_rec,
					n_classes = 2,
					model = "gamma0",
					algorithm = "MLE",
					print_iterations = FALSE)
```

```{r, summary_mdcev_lc}
	SummaryMDCEV(mdcev.lc)
```

In this LC example, we assume that there are two types of people that have different preferences for recreation. The probability of class assignment depends on unobserved factors and the three sociodemographic factors included in the membership equation

### Estimating Random Parameters MDCEV

*too come*

## Rmdcev Demand and Welfare Simulation

The **rmdcev** package includes simulation functions for forecasting demand and calculating welfare measures. I first provide an overview of the approach used for simulation and then provide code examples.

There are two steps to simulation approach. In the first step we draw simulated values for the unobserved heterogeneity term ($\varepsilon$) using Monte Carlo simulation techniques. The second step uses these draws, model parameters and the data to calculate Marhsallian or Hicksian demands. 

### Step 1: Simulating unobervered heterogeneity

Monte Carlo simulation techniques can be employed to draw simulated values of the unobserved heterogeneity ($\varepsilon$) using either unconditional or conditional draws. 

1. Unconditional error draws: draw from the entire distribution of unobserved heterogeneity using the following formula

$\varepsilon_{k} = -log(-log(uniform(0,1)))) * \sigma$

where $uniform(0,1)$ is a uniform random draw and $\sigma$ is the scale parameter.

2. Conditional error draws: draw errors dependent on whether alternative is consumed or not

- If $x_k>0$, set  $\varepsilon_k = (V_1 - V_k)/ \sigma$ where $V_1$ and $V_k$ depend on the model specification as detailed above.
- If $x_k=0$, $\varepsilon_k < (V_1 - V_k)/ \sigma$ and we can simulate $\varepsilon_k$ from the truncated type I extreme value distribution such that 

$\varepsilon_k = -log(-log(uniform(0, 1) * exp(-exp(\frac{v1 - vk}{\sigma})))) * \sigma$

In both cases, we normalize $\varepsilon_1=0$. 

In the conditional approach, we draw errors such that the model perfectly predicts the observed consumption patterns in the baseline state \citep{von_haefen_kuhn-tucker_2005}. Thus the conditional approach uses observed behaviour by individuals to characterize unobserved heterogeneity.

Another difference is more computational as the conditional error draws avoids the need to calculate consumption patterns in the baseline state as well as simulate the entire distribution of unobserved heterogeneity.

**rmdcev** allows errors to be drawn using uniform draws or the Modified Latin Hypercube Sampling algorithm.

### Step 2: Calculating Demand and Welfare measures

With the error draws in hand, the second step is to simulate demand or welfare. **rmdcev** implements the Pinjari and Bhat (2011) demand forecasting routine for simulating demand and the approach described by Lloyd-Smith (2017) for welfare calculations. 

#### Marshallian Demand

Pinjari and Bhat (2011)

```{r, sim_mdcev_demand}
	npols <- 2
	policies <-	CreateBlankPolicies(npols, mdcev.fit$stan_data[["J"]], mdcev.fit$stan_data[["dat_psi"]])
	policies$price_p[[1]] <- c(0, rep(5,mdcev.fit$stan_data[["J"]]))
	policies$price_p[[2]] <- c(0, rep(10,mdcev.fit$stan_data[["J"]]))
	df_sim <- PrepareSimulationData(mdcev.fit, policies)

#	demand <- SimulateMDCEV(df_sim$df_indiv, df_common = df_sim$df_common, sim_options = df_sim$sim_options,
#						 cond_err = 1, nerrs = 20, sim_type = "demand")

```

#### Welfare Analysis

```{r, sim_mdcev_welfare_setup}
	npols <- 2
	policies <-	CreateBlankPolicies(npols, mdcev.fit$stan_data[["J"]], mdcev.fit$stan_data[["dat_psi"]])
	policies$price_p[[1]] <- c(0, rep(5,mdcev.fit$stan_data[["J"]]))
	policies$price_p[[2]] <- c(0, rep(10,mdcev.fit$stan_data[["J"]]))
	df_sim <- PrepareSimulationData(mdcev.fit, policies)
```


```{r, sim_mdcev_welfare}
	wtp <- SimulateMDCEV(df_sim$df_indiv, df_common = df_sim$df_common, sim_options = df_sim$sim_options,
						 cond_err = 1, nerrs = 15, sim_type = "welfare")
	SummaryWelfare(wtp)
```

Lloyd-Smith (2017)

## Modelling notes



## Conclusions

The **rmdcev** package implements the MDCEV . In this paper, we have demonstrated the use of the pacakage to estimate several model specifications and to derive welfare.

## Appendix



### Specific steps in algorithm for hybrid-profile utility functions

**Step 0**: Assume that only the numeraire alternative is chosen and let the number of chosen alternatives equal one (M=1).

**Step 1**: Using the data, model parameters, and either conditional or unconditional simulated error term draws, calculate the price-normalized baseline utility values ($\psi_k/p_k$) for all alternatives. Sort the $K$ alternatives in the descending order of their price-normalized baseline utility values. Note that the numeraire alternative is in the first place. Go to step 2.

**Step 2**: Compute the value of $\lambda^E$ using the following equation:

\begin{align}
\label{lambdaE}
\frac{1}{\lambda^E} = \left[ \frac{\alpha \bar{U} + \sum_{m=2}^{M} \gamma_m \psi_m} {\sum_{m=2}^{M} \gamma_m \psi_m \left( \frac{p_m}{\psi_m} \right)^\frac{\alpha}{\alpha-1} + \psi_1 \left(\frac{p_1}{\psi_1} \right)^\frac{\alpha}{\alpha-1}} \right] ^\frac{\alpha-1}{\alpha}.
\end{align}

Go to step 3.

**Step 3**: If $\frac{1}{\lambda^E} > \frac{\psi_{M+1}}{p_{M+1}}$, go to step 4. Else if $\frac{1}{\lambda^E} < \frac{\psi_{M+1}}{p_{M+1}}$, set $M = M + 1$. If $M < K$, go back to step 2. If $M = K$, go to step 4.

**Step 4**: Compute the optimal Hicksian consumption levels for the first $I$ alternatives in the above descending order using the following equations

\begin{align}
\begin{split}
\label{foc13}
x_1 &=   \left( \frac{p_1}{\lambda^E \psi_1} \right)^\frac{1}{\alpha_1-1}\text{, and} \\
x_m &=   \left[ \left( \frac{p_m}{\lambda^E \psi_m} \right)^\frac{1}{\alpha_m-1}-1 \right]\gamma_m, \; \; \text{if} \; \; x_m > 0.
\end{split}
\end{align}

Set the remaining alternative consumption levels to zero and stop.

### Specific steps in algorithm for general utility functions:

In the more general case of a utility function with different $\alpha_k$ parameters for each alternative, I modify the algorithm in Section 3.2. In this context, there is no closed-form expressions for $\lambda^E$ and I need to conduct a numerical bisection routine. Let $\hat{\lambda^E}$ and $\hat{U}$ be estimates of $\lambda^E$ and $U$ and let $tol_{\lambda}$ and $tol_{U}$ be the tolerance levels for estimating $\lambda^E$ and $U$ that can be arbitrarily small.

**Step 0**: Assume that only the numeraire is chosen and let the number of chosen alternatives equal one (M=1).

**Step 1**: Using the data, model parameters, and either conditional or unconditional simulated error term draws, calculate the price-normalized baseline utility values ($\psi_k/p_k$) for all alternatives. Sort the $K$ alternatives in the descending order of their price-normalized baseline utility values. Note that the numeraire is in the first place. Go to step 2.

**Step 2**: Let $\frac{1}{\hat{\lambda^E}} = \frac{\psi_{M+1}}{p_{M+1}}$ and substitute $\hat{\lambda^E}$ into the following quation to obtain an estimate of $\hat{U}$.

\begin{align}
\bar{U}=\sum_{M=2}^{M} \frac{\gamma_m}{\alpha_m}\psi_m \left[ \left( \frac{p_m}{\lambda^E \psi_m} \right)^\frac{\alpha_m}{\alpha_m-1} - 1 \right] + \frac{\psi_1}{\alpha_1}\left(\frac{p_1}{\lambda^E \psi_1} \right)^\frac{\alpha_1}{\alpha_1-1}.
\label{util_algo}
\end{align}

**Step 3**: If $\hat{U} < \bar{U}$, go to step 4. Else, if $\hat{U} \geq \bar{U}$, set $\frac{1}{\lambda_l^E}= \frac{\psi_{M+1}}{p_{M+1}}$ and $\frac{1}{\lambda_u^E}= \frac{\psi_{M}}{p_{M}}$. Go to step 5.

**Step 4**: Set $M=M+1$. If $M<K$, go to step 2. Else if $M=K$, set $\frac{1}{\lambda_l^E}= 0$ and $\frac{1}{\lambda_u^E}= \frac{\psi_{K}}{p_{K}}$. Go to step 5.

**Step 5**: Let $\hat{\lambda^E}= (\lambda_l^E+\lambda_u^E)/2$ and substitute $\hat{\lambda^E}$ into Equation (\ref{util_algo}) to obtain an estimate of $\hat{U}$. Go to step 6.

**Step 6**: If $|\lambda_l^E-\lambda_u^E| \leq \; tol_{\lambda}$ or $|\hat{U}-\bar{U}| \leq \; tol_{U}$, go to step 7. Else if $\hat{U}<\bar{U}$, update $\lambda^E_u= (\lambda_l^E+\lambda_u^E)/2$ and go to step 5. Else if $\hat{U}>\bar{U}$, update $\lambda^E_l= (\lambda_l^E+\lambda_u^E)/2$ and go to step 5.

**Step 7**: Compute the optimal Hicksian consumption levels for the first $I$ alternatives in the above descending order using Equation (\ref{foc13}). Set the remaining alternative consumption levels to zero and stop.
